### 2016-06-25 store-consistency.tex

In practice, relational databases resolve the conflict by prioritizing
weak isolation (thus, concurrency and performance) over strong
consistency, so the execution traces do not necessarily satisfy {\sc
sc}. In particular, visibility constraints imposed by {\sc sc} are
violated if (and only if) they are found to be in conflict with the
constraints imposed by the isolation level. In the context of
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store makes $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term
\emph{recency}~\cite{bailisvldb} is often used in place of consistency
to capture the best-effort nature of the {\sc sc} guarantee on
databases.}. We generalize this approach to any $\I_s$ and $\I_c$ by
defining a \emph{maximum visiblity principle} to determine an
acceptable weakening of $\I_s$ in case of a conflict with $\I_c$.  The
principle requires the weakened consistency guarantee ($\I_s'$) of the
store to enforce all visibility relationships imposed by the actual
consistency guarantee ($\I_s$), unless enforcing such a relationship
violates $\I_c$. Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}

Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes a practical {\sc sc} store, such as a relational database.

@@

\subsection{Eventually Consistent Replication}

@@

In this section, we present an intuitive explanation of the soundness
result in context of eventually consistent replication. 
% In other words, we explain why our approach is a sound way of
% reasoning about transactions under eventually consistent replication
% despite our system model not including the standard artifacts
% associated with a replicated store.

Existing approaches to reason about eventually consistent replication
necessarily involve reasoning in terms of \emph{replicas} of data. The
primary challenge in this setting is to ensure that the assumptions
made and guarantees enforced by an operation at one replica carry over
to other replicas that merge its effects, thus preserving the overall
integrity of the system. Reasoning frameworks, such
as~\cite{gotsmanpopl16}, address this challenge by imposing
restrictions on how various replica states differ, i.e., by
strengthening the consistency of the store. Our view of eventually
consistent replication however does not explicitly involve replicas.
Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of EC replication
with the conventional replica-based model.  Under our model, the
notion of a replica is subsumed by the concept of visibility; a
replica is defined by the subset ($S$) of global state ($\E.\A$) that
an operation witnesses.  Constraints over replica states therefore
manifest as constraints over the visibility relation. For example,
instead of requiring the store to be causally consistent, an operation
can request to witness a causally consistent subset of the state; such
demands can be made via the trace invariant $\I$. For a precondition
($P$) of the operation to be useful, it has to be an assertion over
every causally consistent subset of the global state. Since any
replica that eventually executes the operation has to expose one such
subset ($S$), the precondition is guaranteed to hold regardless of the
replica. There is however one problem with this explanation; by
considering subsets of just one global state, it ignores the fact that
the global state (hence, the replica states) change during the
execution of the operation. Existing approaches account for this
change by distinguishing between \emph{effect generation} event at one
replica and \emph{effect merge} event at another replica, and
requiring that certain invariants be preserved between these two
events at different replicas. Our framework folds all of this into a
stability condition (\S\ref{sec:rely-guarantee}). Since any change to
the global state during the execution of the operation is an
interference, and $P$ is required to be stable with respect to any
such interference, it follows that $P$ is valid on every replica, thus
ensuring that assumptions made at the generation event are also valid
at the merge event.

>>>>

In practice, relational databases resolve the conflict by prioritizing
weak isolation (thus, concurrency and performance) over strong
consistency, so the execution traces do not necessarily satisfy {\sc
sc}. In particular, visibility constraints imposed by {\sc sc} are
violated if (and only if) they are found to be in conflict with the
constraints imposed by the isolation level. In the context of
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store needs to make $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term
\emph{recency}~\cite{bailisvldb} is often used in place of consistency
to capture the best-effort nature of the {\sc sc} guarantee on
databases.}. We generalize this approach to any $\I_s$ and $\I_c$ by
defining a \emph{maximum visiblity principle} to determine an
acceptable weakening of $\I_s$ in case of a conflict with $\I_c$.  The
principle requires the weakened consistency guarantee ($\I_s'$) of the
store to enforce all visibility relationships imposed by the actual
consistency guarantee ($\I_s$), unless enforcing such a relationship
violates $\I_c$. 
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes relational database-like stores.

@@

\subsection{Weakly Consistent Replication}

@@

Reasoning under weakly consistent replication has received special
attention in the recent research~\cite{gotsmanpopl16}. Our operational
semantics and proof system are however more general and admit
replication as a special case. In this section, we explain how our
reasoning approach works under weakly consistent replication by
comparing our model with the standard model of replication.

Reasoning about eventually consistent replication often involves
reasoning in terms of \emph{replicas} of data. The primary challenge
in this setting is to ensure that the assumptions made and guarantees
enforced by an operation at one replica carry over to other replicas
that merge its effects, thus preserving the overall integrity of the
system. In~\cite{gotsmanpopl16}, this challenge is partly addressed by
imposing restrictions on how various replica states differ, i.e., by
fixing a system model with a stronger baseline consistency ({\sc cc})
than {\sc ec}. This unfortunately restricts the applicability of
reasoning approach from being extended to weaker stores such as
Bayou~\cite{bayou} or Quelea~\cite{pldi15}. Our view of replication
however does not explicitly involve replicas.
Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of weakly
consistent replication with the conventional replica-based model.
Under our model, the notion of a replica is subsumed by the concept of
visibility; a replica is defined by the subset ($S$) of global state
($\E.\A$) that an operation witnesses. Constraints over replica states
therefore manifest as constraints over the visibility relation.  For
example, instead of requiring the store to be causally consistent, an
operation can request to witness a causally consistent subset of the
state; such demands can be made via the trace invariant $\I$. For a
precondition ($P$) of the operation to be useful, it has to be an
assertion over every causally consistent subset of the global state.
Since any replica that eventually executes the operation has to expose
one such subset ($S$), the precondition is guaranteed to hold
regardless of the replica. There is however one problem with this
explanation; by considering subsets of just one global state, it
ignores the fact that the global state (hence, the replica states)
change during the execution of the operation. To account for this
change,~\cite{gotsmanpopl16} distinguishes between \emph{effect
generation} event at one replica and \emph{effect merge} event at
another replica, and requires that certain invariants be preserved
between these two events at different replicas. Our framework folds
all of this into a stability condition (\S\ref{sec:rely-guarantee}).
Since any change to the global state during the execution of the
operation is an interference, and $P$ is required to be stable with
respect to any such interference, it follows that $P$ is valid on
every replica, thus ensuring that assumptions made at the generation
event are also valid at the merge event.
